Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2011-10-18 boettiger.tex
bib changes to follow editorial constraints

                             %% Enter the text of your article beginning here and ending before
                             %% \begin{acknowledgements}
                             %% Section head commands for your reference:
                             %% \section{}
                             %% \subsection{}
                             %% \subsubsection{}

\dropcap{T}here is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\cite{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠. 
Many important ecosystems may currently be threatened with collapse, including corals~\cite{Bellwood2004}, fisheries~\cite{Berkes2006}⁠, lakes~\cite{Carpenter2011}, and semi-arid ecosystems~\cite{Kefi2007}⁠. 
Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
and the idea that management responses would be needed either to avoid an undesirable shift or else to adapt to novel conditions,
it is important to develop the ability to predict impending regime shifts based on early warning signs. 
With a good model of the system, detail-oriented approaches could be useful~\cite{Lenton2009},
but in cases where specific models are not available general approaches are needed~\cite{Scheffer2009}⁠.
We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system.\footnote{A system involving a local bifurcation where the eigenvalue approaches zero \cite{Scheffer2001, Scheffer2009}.}  
Making this assumption provides a ``best-case'' test of detection approaches;
we will demonstrate the resulting uncertainty can be very high.
We describe a novel statistical approach for handling uncertainty that is appropriate for the goal of informing management.

@@

\section{Examples}

@@

                             %% Enter the largest bibliography number in the facing curly brackets
                             %% following \begin{thebibliography}

                             \bibliographystyle{/home/cboettig/Documents/tex/pnas.bst}
                             \bibliography{/home/cboettig/Documents/bibliographies/library.bib}
                           \end{article}
                           %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

                           %% For figures, put the caption below the illustration.
                           %%
                           %% \begin{figure}
                           %% \caption{Almost Sharp Front}\label{afoto}
                           %% \end{figure}
 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig1}
     \label{fig1}
     \caption{Top row: The distributions of a hypothetical warning indicator are shown under the case of a stable system (blue) and a system approaching a critical transition (red).  Bottom row: Points along the ROC curve are calculated for each possible threshold indicated in the top row.  The false positive rate is the integral of the distribution of the test statistic under the stable system right of the threshold (blue shaded area, corresponding to blue vertical line).  The true positive rate is the integral of the system approaching a transition left of the threshold (red shaded area, corresponds to the red line).  Successive columns show the threshold increasing, tracing out the ROC curve (successively gray lines show the threshold used for the matching gray point added to the ROC curve).}
  \end{center}
 \end{figure*}

 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig2}
     \label{fig2}
     \caption{Early warning signals in simulated and empirical data sets.  The first two columns are simulated data from (a) a stable system (Stable), and (b) the same system approaching a saddle-node bifurcation due to a gradually deteriorating environment (Deteriorating).  Empirical examples are from (c) \emph{Daphnia magna} concentrations manipulated in a chemostat towards a critical transition (Daphnia), and (d) deuterium concentrations previously cited as an early warning signal of a critical transition in glaciation (Glaciation). Increases in summary statistics such as variance, autocorrelation, or skew, computed over a moving window, are used to indicate if a system is moving towards a critical transition (see supplementary online text).  The increase is measured by the correlation coefficient $\tau$.  Note that positive correlation does not guarantee the system is moving towards a transition, as seen in the stable system, first column.}
  \end{center}
 \end{figure*}

 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig3.pdf}
     \label{fig3}
     \caption{ROC curves for different early warning indicators on simulated and empirical data.  The area under the curve, inset, indicates how the method compares to a random guess (0.5).  The likelihood method performs substantially better than other metrics, particularly on the weaker trends seen in the empirical data. The simulated data sets have 40 points, ``Daphnia'' has 16, the ``Glaciation'' has 121. See corresponding distributions, Fig. S1.}
  \end{center}
 \end{figure*}

 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig4.pdf}
     \label{fig4}
     \caption{ROC curves (from Fig. 3) are recalculated under different sampling frequencies in the simulation step (see supplementary online text), using a total of 25, 50, or 200 points.  Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective. A warning sighal is not detected in stable sytem simulation. Corresponding distributions shown in Fig S2-5.} 
  \end{center}
 \end{figure*}

>>>>

                             %% Enter the text of your article beginning here and ending before
                             %% \begin{acknowledgements}
                             %% Section head commands for your reference:
                             %% \section{}
                             %% \subsection{}
                             %% \subsubsection{}
                             \section{Introduction}
\dropcap{T}here is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\cite{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠. 
Many important ecosystems may currently be threatened with collapse, including corals~\cite{Bellwood2004}, fisheries~\cite{Berkes2006}⁠, lakes~\cite{Carpenter2011}, and semi-arid ecosystems~\cite{Kefi2007}⁠. 
Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
and the idea that management responses would be needed either to avoid an undesirable shift or else to adapt to novel conditions,
it is important to develop the ability to predict impending regime shifts based on early warning signs. 
With a good model of the system, detail-oriented approaches could be useful~\cite{Lenton2009},
but in cases where specific models are not available general approaches are needed~\cite{Scheffer2009}⁠.
We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system.\footnote{A system involving a local bifurcation where the eigenvalue approaches zero \cite{Scheffer2001, Scheffer2009}.}  
Making this assumption provides a ``best-case'' test of detection approaches;
we will demonstrate the resulting uncertainty can be very high.
We describe a novel statistical approach for handling uncertainty that is appropriate for the goal of informing management.

@@

\section{Results}

@@

                             %% Enter the largest bibliography number in the facing curly brackets
                             %% following \begin{thebibliography}
\begin{thebibliography}{10}

\bibitem{Holling1973}
Holling CS
\newblock (1973) {Resilience and Stability of Ecological Systems}.
\newblock \emph{Annual Review of Ecology and Systematics} 4:1--23.

\bibitem{Wissel1984}
Wissel C
\newblock (1984) {A universal law of the characteristic return time near
  thresholds}.
\newblock \emph{Oecologia} 65:101--107.

\bibitem{Scheffer2001}
Scheffer M, Carpenter SR, Foley JA, Folke C, Walker B
\newblock (2001) {Catastrophic shifts in ecosystems.}
\newblock \emph{Nature} 413:591--6.

\bibitem{Scheffer2009}
Scheffer M, {et~al.}
\newblock (2009) {Early-warning signals for critical transitions.}
\newblock \emph{Nature} 461:53--9.

\bibitem{Drake2010}
Drake JM, Griffen BD
\newblock (2010) {Early warning signals of extinction in deteriorating
  environments}.
\newblock \emph{Nature} 467:456--459.

\bibitem{Carpenter2011}
Carpenter SR, {et~al.}
\newblock (2011) {Early Warnings of Regime Shifts: A Whole-Ecosystem
  Experiment}.
\newblock \emph{Science} 1079.

\bibitem{Bellwood2004}
Bellwood DR, Hughes TP, Folke C, Nystr\"{o}m M
\newblock (2004) {Confronting the coral reef crisis.}
\newblock \emph{Nature} 429:827--33.

\bibitem{Berkes2006}
Berkes F, {et~al.}
\newblock (2006) {Ecology. Globalization, roving bandits, and marine
  resources.}
\newblock \emph{Science (New York, N.Y.)} 311:1557--8.

\bibitem{Kefi2007}
K\'{e}fi S, {et~al.}
\newblock (2007) {Spatial vegetation patterns and imminent desertification in
  Mediterranean arid ecosystems.}
\newblock \emph{Nature} 449:213--7.

\bibitem{Lenton2009}
Lenton TM, {et~al.}
\newblock (2009) {Using GENIE to study a tipping point in the climate system.}
\newblock \emph{Philosophical transactions. Series A, Mathematical, physical,
  and engineering sciences} 367:871--84.

\bibitem{Carpenter2006}
Carpenter SR, Brock WA
\newblock (2006) {Rising variance: a leading indicator of ecological
  transition.}
\newblock \emph{Ecology letters} 9:311--8.

\bibitem{Dakos2008}
Dakos V, {et~al.}
\newblock (2008) {Slowing down as an early warning signal for abrupt climate
  change.}
\newblock \emph{Proceedings of the National Academy of Sciences of the United
  States of America} 105:14308--12.

\bibitem{Guttal2008}
Guttal V, Jayaprakash C
\newblock (2008) {Spatial variance and spatial skewness: leading indicators of
  regime shifts in spatial ecological systems}.
\newblock \emph{Theoretical Ecology} 2:3--12.

\bibitem{Guttal2008a}
Guttal V, Jayaprakash C
\newblock (2008) {Changing skewness: an early warning signal of regime shifts
  in ecosystems.}
\newblock \emph{Ecology letters} 11:450--60.

\bibitem{Dakos2011}
Dakos V, K\'{e}fi S, Rietkerk M, Nes EHV, Scheffer M
\newblock (2011) {Slowing Down in Spatially Patterned Ecosystems at the Brink
  of Collapse}.
\newblock \emph{The American Naturalist}.

\bibitem{Contamin2009}
Contamin R, Ellison AM
\newblock (2009) {Indicators of regime shifts in ecological systems: what do we
  need to know and when do we need to know it?}
\newblock \emph{Ecological applications : a publication of the Ecological
  Society of America} 19:799--816.

\bibitem{Scheffer2010}
Scheffer M
\newblock (2010) {Complex systems: Foreseeing tipping points.}
\newblock \emph{Nature} 467:411--2.

\bibitem{Inman2011}
Inman M
\newblock (2011) {Sending out an SOS}.
\newblock \emph{Nature Climate Change} pp 1--4.

\bibitem{Green1989}
Green DM, Swets JA
\newblock (1989) \emph{{Signal Detection Theory and Psychophysics}}
\newblock (Peninsula Pub), p 521.

\bibitem{Keller2009}
Keller RP, Lodge DM, Lewis MA, Shogren JF, eds
\newblock (2009) \emph{{Bioeconomics of Invasive Species: Integrating Ecology,
  Economics, Policy, and Management}}
\newblock (Oxford University Press, USA), p 320.

\bibitem{Dakos2009}
Dakos V, Nes EH, Donangelo R, Fort H, Scheffer M
\newblock (2009) {Spatial correlation as leading indicator of catastrophic
  shifts}.
\newblock \emph{Theoretical Ecology} pp 163--174.

\bibitem{VanNes2007}
van Nes EH, Scheffer M
\newblock (2007) {Slow recovery from perturbations as a generic indicator of a
  nearby catastrophic shift.}
\newblock \emph{The American Naturalist} 169:738--47.

\bibitem{Biggs2009}
Biggs R, Carpenter SR, Brock Wa
\newblock (2009) {Turning back from the brink: detecting an impending regime
  shift in time to avert it.}
\newblock \emph{Proceedings of the National Academy of Sciences of the United
  States of America} 106:826--31.

\bibitem{Cox1961}
Cox DR
\newblock (1961) \emph{{Tests of Seperate Families of Hypotheses}}
\newblock No.{}~2, pp 105 -- 123.

\bibitem{Hastings2010}
Hastings A, Wysham DB
\newblock (2010) {Regime shifts in ecological systems can occur with no
  warning.}
\newblock \emph{Ecology letters}.

\bibitem{Schreiber2003}
Schreiber SJ
\newblock (2003) {Allee effects, extinctions, and chaotic transients in simple
  population models}.
\newblock \emph{Theoretical population biology} 64:201--209.

\bibitem{Schreiber2008}
Schreiber SJ, Rudolf VHW
\newblock (2008) {Crossing habitat boundaries: coupling dynamics of ecosystems
  through complex life cycles.}
\newblock \emph{Ecology letters} 11:576--87.

\bibitem{Huelsenbeck1996}
Huelsenbeck JP, Bull JJ
\newblock (1996) {A Likelihood Ratio Test to Detect Conflicting Phylogenetic
  Signal}.
\newblock \emph{Systematic Biology} 45:92--98.

\bibitem{Petit1999}
Petit JR, {et~al.}
\newblock (1999) {Climate and atmospheric history of the past 420,000 years
  from the Vostok ice core, Antarctica}.
\newblock \emph{Nature} 399:429--436.

\bibitem{Strogatz2001a}
Strogatz SH
\newblock (2001) \emph{{Nonlinear Dynamics And Chaos: With Applications To
  Physics, Biology, Chemistry, And Engineering (Studies in Nonlinearity)}}
\newblock (Westview Press), p 512.

\bibitem{Guckenheimer1983}
Guckenheimer J, Holmes P
\newblock (1983) \emph{{Nonlinear Oscillations, Dynamical Systems, and
  Bifurcations of Vector Fields (Applied Mathematical Sciences Vol. 42)}}
\newblock (Springer), p 480.

\bibitem{Levins1969}
Levins R
\newblock (1969) {Some demographic and genetic consequences of environmental
  heterogeneity for biological control}.
\newblock \emph{Bulletin of the Entomological Society of America} 15:237--240.

\bibitem{Gardiner2009}
Gardiner C
\newblock (2009) \emph{{Stochastic Methods: A Handbook for the Natural and
  Social Sciences (Springer Series in Synergetics)}}
\newblock (Springer), p 447.

\end{thebibliography}
                           %  \bibliographystyle{/home/cboettig/Documents/tex/pnas.bst}
                            % \bibliography{/home/cboettig/Documents/bibliographies/library.bib}
                           \end{article}
                           %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

                           %% For figures, put the caption below the illustration.
                           %%
                           %% \begin{figure}
                           %% \caption{Almost Sharp Front}\label{afoto}
                           %% \end{figure}
                           \section{Figure Legends}
\begin{figure}[hb]
   \begin{center}
%     \includegraphics[width=\linewidth]{Fig1}
     \label{fig1}
     \caption{Top row: The distributions of a hypothetical warning indicator are shown under the case of a stable system (blue) and a system approaching a critical transition (red).  Bottom row: Points along the ROC curve are calculated for each possible threshold indicated in the top row.  The false positive rate is the integral of the distribution of the test statistic under the stable system right of the threshold (blue shaded area, corresponding to blue vertical line).  The true positive rate is the integral of the system approaching a transition left of the threshold (red shaded area, corresponds to the red line).  Successive columns show the threshold increasing, tracing out the ROC curve (successively gray lines show the threshold used for the matching gray point added to the ROC curve).}
  \end{center}
 \end{figure}

 \begin{figure}[h]
   \begin{center}
%     \includegraphics[width=\linewidth]{Fig2}
     \label{fig2}
     \caption{Early warning signals in simulated and empirical data sets.  The first two columns are simulated data from (a) a stable system (Stable), and (b) the same system approaching a saddle-node bifurcation due to a gradually deteriorating environment (Deteriorating).  Empirical examples are from (c) \emph{Daphnia magna} concentrations manipulated in a chemostat towards a critical transition (Daphnia), and (d) deuterium concentrations previously cited as an early warning signal of a critical transition in glaciation (Glaciation). Increases in summary statistics such as variance, autocorrelation, or skew, computed over a moving window, are used to indicate if a system is moving towards a critical transition (see supplementary online text).  The increase is measured by the correlation coefficient $\tau$.  Note that positive correlation does not guarantee the system is moving towards a transition, as seen in the stable system, first column.}
  \end{center}
 \end{figure}

 \begin{figure}[h]
   \begin{center}
%     \includegraphics[width=\linewidth]{Fig3.pdf}
     \label{fig3}
     \caption{ROC curves for different early warning indicators on simulated and empirical data.  The area under the curve, inset, indicates how the method compares to a random guess (0.5).  The likelihood method performs substantially better than other metrics, particularly on the weaker trends seen in the empirical data. The simulated data sets have 40 points, ``Daphnia'' has 16, the ``Glaciation'' has 121. See corresponding distributions, Fig. S1.}
  \end{center}
 \end{figure}

 \begin{figure}[h!]
   \begin{center}
%     \includegraphics[width=\linewidth]{Fig4.pdf}
     \label{fig4}
     \caption{ROC curves (from Fig. 3) are recalculated under different sampling frequencies in the simulation step (see supplementary online text), using a total of 25, 50, or 200 points.  Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective. A warning sighal is not detected in stable sytem simulation. Corresponding distributions shown in Fig S2-5.} 
  \end{center}
 \end{figure}
